Gammapy: A Python package for gamma-ray astronomy

### 2023-03-06 src/ms-review.tex
Add code stats file and improve plot
Fix #110
Use siunitx package

\usepackage{graphicx}
\usepackage{url}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{import}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage[autostyle]{csquotes}

@@

%% \gammaray astronomy
%% CN: I think we should use the expression Water Cherenkov Detectors and use the
%% abbreviation (WCD) all along the text, later we use "particle samplers"
The \gammaray range of the electromagnetic spectrum provides us insights into the
most energetic processes in the universe such as those accelerating particles in the surroundings of
black holes, and remnants of supernova explosions. As in other
branches of astronomy, \gammarays can be observed by
satellite as well as ground based instruments.
Ground-based instruments use the Earth's atmosphere as a particle detector.
Very-high-energy (VHE) cosmic \gammarays interact in the atmosphere and
create large showers of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on these extensive air showers to detect the
primary \gammaray photons and infer their incident direction and energy.
VHE \gammaray astronomy covers the energy range from few tens of ${\rm GeV}$ up to the ${\rm PeV}$.
There are two main categories of ground-based instruments:

@@

\section{Gamma-ray Data Analysis}
\begin{figure*}[h!]
	\centering
	\includegraphics[width=1.\textwidth]{figures/data_flow.pdf}
	\caption{
		\gammapy sub-package structure and data analysis workflow. The top row
        defines the different levels of data reduction, from lists of \gammaray-like
        events on the left (DL3), to high-level scientific products products
        (DL5) on the right. The direction of the data flow is illustrated with the
        gray arrows. The gray folder icons represent the different sub-packages
        in \gammapy and their names. Below each icon there is a list of the most
        important objects defined in the sub-package.
    }
	\label{fig:data_flow}
\end{figure*}
%
\label{sec:gammaray-data-analysis}
% This might not be the bast place compared to introduction
The data analysis process in \gammaray astronomy is usually split into two parts.
The first one deals with the data processing from camera measurement, calibration, event
reconstruction and selection to yield a list of reconstructed \gammaray event candidates.
This part of the data reduction sequence, sometimes referred to as low-level analysis,
is usually very specific to a given observation technique and even to a given instrument.

@@

To extract physically relevant information, such as the flux, spatial or spectral shape of one or more sources,
an analytical model is commonly adopted to describe the intensity of the radiation from gamma-ray sources
as a function of the energy, $E_{\rm true}$, and of the position in the FoV, $p_{\rm true}$:
\begin{equation}
    \Phi(p_{\rm true}, E_{\rm true}; \hat{\theta}) \quad \mathrm{[TeV^{-1}cm^{-2}s^{-1}]}
    \label{eq:general_source_model}
\end{equation}
where $\hat{\theta}$ is a set of model parameters that can be adjusted in a fit. To convert this analytical flux model
into a prediction on the number of gamma-ray events detected by an instrument, $N_{\rm pred}$,
with their estimated energy $E$ and position $p$, the model is convolved with the response function of the instrument.

@@

\subsection{gammapy.irf}
\label{ssec:gammapy-irf}
%
%
\begin{figure*}[ht!]
	\centering
	\includegraphics[width=1.\textwidth]{figures/irfs.pdf}
	\caption{
		Using \code{gammapy.irf} to read and plot instrument response functions.
		The left panel shows the effective area as a function of energy for
		the \cta, \hess, \magic, \hawc and \fermi instruments. The right panel shows
		the $68\%$ containment radius of the PSF as a function of energy for the \cta, \hess
		and \fermi instruments. The \cta IRFs are from the \textit{prod5} production. The \hess IRFs are from the DL3 DR1,
        using observation ID 033787. The \magic effective area is computed for a
        $20\,{\rm min}$ observation at the Crab Nebula coordinates. The
		\fermi IRFs use \textit{pass8} data and are also taken at the position of the Crab Nebula.
		The \hawc effective area is shown for the event classes $N_{Hit}=5 - 9$ as light gray
		lines along with the sum of all event classes as a black line. The \hawc IRFs are taken from
		the first public release of event data the \hawc collaboration. All IRFs do not correspond
		to the latest performance of the instruments, but still are representative of the 
		detector type and energy range. We exclusively relied on publicly available
		data provided by the collaborations. The data is also availabe in the
		\code{gammapy-data} repository.
    }
	\label{fig:irfs}
\end{figure*}
%

@@

\begin{table*}
    \begin{center}
        \begin{tabular}{lll}
         \hline
         Type & Description & Unit Equivalency \\
         \hline
         dnde & Differential flux at a given energy & $\mathrm{TeV^{-1}~cm^{-2}~s^{-1}}$ \\
         e2dnde & Differential flux at a given energy  & $\mathrm{TeV~cm^{-2}~s^{-1}}$ \\
         flux & Integrated flux in a given energy range & $\mathrm{cm^{-2}~s^{-1}}$ \\
         eflux & Integrated energy flux in a given energy range & $\mathrm{erg~cm^{-2}~s^{-1}}$\\
         \hline
        \end{tabular}
    \end{center}
    \caption{Definition of the different SED types supported in \gammapy.}
    \label{tab:sed_types}
\end{table*}

@@

\subsection{Temporal Analysis}
\label{ssec:temporal-analysis}
A common use case in many astrophysical scenarios is to study the temporal
variability of a source. The most basic way to do this is to construct a
light curve, i.e., the flux of a source in each given time bin. In \gammapy, this
is done by using the \code{LightCurveEstimator} that fits the normalisation of a
source in each time (and optionally energy) band per observation, keeping constant
other parameters.
For custom time binning, an observation needs to be split into finer time bins using
the \code{Observation.select\_time} method. Figure~\ref{fig:hess_lightcurve_pks}
shows the light curve of the blazar PKS~2155-304 in different energy bands as
observed by the \hess telescope during an exceptional flare on the night of
July 29 - 30, 2006~\cite{2009A&A...502..749A}. The data are publicly available 
as a part of the HESS-DL3-DR1~\cite{HESS_DR1}. Each observation is first split into 10 min smaller
observations, and spectra extracted for each of these within a 0.11\textdegree~radius
around the source. A \code{PowerLawSpectralModel} is fit to all the datasets, leading
to a reconstructed index of $3.54 \pm 0.02$. With this adjusted spectral model
the \code{LightCurveEstimator} runs directly for two energy bands, $0.5\,{\rm TeV} - 1.5\,{\rm TeV}$,
and $1.5\,{\rm TeV} - 20\,{\rm TeV}$, respectively.
%
\begin{figure*}[t]
    \sidecaption
	\includegraphics[width=0.6666\textwidth]{figures/hess_lightcurve_pks.pdf}
	\caption{
        Binned light curves in two different energy bands for the source
        PKS~2155-304 in two energy bands ($0.5\,{\rm TeV} - 1.5\,{\rm TeV}$, and $1.5\,{\rm TeV} - 20\,{\rm TeV}$)
        as observed by the \hess telescopes in 2006. The coloured markers
        show the flux points in the different energy bands. The horizontal
        error illustrates the width of the time bin of 10~min. The vertical
        error bars show the associated asymmetrical flux errors. The marker
        is set to the center of the time bin.
    }
    \label{fig:hess_lightcurve_pks}
\end{figure*}
%
The obtained flux points can be analytically modelled using the available or
user-implemented temporal models. Alternatively, instead of  extracting a
light curve, it is also possible to directly fit temporal models to the reduced
datasets. By associating an appropriate \code{SkyModel}, consisting of both temporal
and spectral components, or using custom temporal models with spectroscopic
variability, to each dataset, a joint fit across the datasets will directly
return the best fit temporal and spectral parameters.

@@

The \fermi data is introduced at the data level DL4, and directly bundled in a
\code{MapDataset}. They have been prepared using the standard \textit{fermitools} \citep{Fermitools2019} and
selecting a region of $5^{\circ}\,\times\,4^{\circ}$ around the
position of the Crab Nebula, applying the same selection criteria of the 3FHL
catalog (7 years of data with energy from $10\,{\rm GeV}$ to $2\,{\rm TeV}$,
~\citealt{3FHL}).

The \magic data is included from the data level DL3. They consist of two
observations of $20\,{\rm min}$ each, chosen from the dataset used to estimate
the performance of the upgraded stereo system~\citep{magic_performance} and
already included in~\cite{joint_crab}. The observations were taken at small
zenith angles ($<30^{\circ}$) in wobble mode~\citep{fomin_1994}, with the
source sitting at an offset of $0.4^{\circ}$ from the FoV center. Their energy
range spans $80\,{\rm GeV} - 20\,{\rm TeV}$. The data reduction for the 1D analysis
is applied, and the data are reduced to a \code{SpectrumDataset} before being fitted.

\hawc data are directly provided as flux points (DL5 data level) and are read
via Gammapy's \code{FluxPoints} class. They were estimated in ~\cite{hawc_crab_2019}
with $2.5\,{\rm years}$ of data and span an energy range $300\,{\rm GeV} - 300\,{\rm TeV}$.

Combining the datasets in a \code{Datasets} list, \gammapy automatically generates
a likelihood including three different types of terms, two Poissonian likelihoods
for \fermi's \code{MapDataset} and MAGIC's \code{SpectrumDataset}, and a $\chi^2$
accounting for the \hawc flux points. For \fermi, a three-dimensional forward folding
of the sky model with the IRF is performed, in order to compute the predicted counts
in each sky-coordinate and energy bin. For \magic, a one-dimensional forward-folding
of the spectral model with the \irfs is performed to predict the counts in each estimated energy bin. A log
parabola is fitted to the almost five decades in energy $10\,{\rm GeV} - 300\,{\rm TeV}$.
%% CN: is the LP formula specified elsewhere?

>>>>

\usepackage{graphicx}
\usepackage{url}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{import}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage[autostyle]{csquotes}
\usepackage{siunitx}

@@

%% \gammaray astronomy
%% CN: I think we should use the expression Water Cherenkov Detectors and use the
%% abbreviation (WCD) all along the text, later we use "particle samplers"
The \gammaray range of the electromagnetic spectrum provides us insights into the
most energetic processes in the universe such as those accelerating particles in the surroundings of
black holes, and remnants of supernova explosions. As in other
branches of astronomy, \gammarays can be observed by
satellite as well as ground based instruments.
Ground-based instruments use the Earth's atmosphere as a particle detector.
Very-high-energy (VHE) cosmic \gammarays interact in the atmosphere and
create large showers of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on these extensive air showers to detect the
primary \gammaray photons and infer their incident direction and energy.
VHE \gammaray astronomy covers the energy range from few tens of \unit{GeV} up to the \unit{PeV}.
There are two main categories of ground-based instruments:

@@

\section{Gamma-ray Data Analysis}
\begin{figure*}[t]
	\centering
	\includegraphics[width=1.\textwidth]{figures/data_flow.pdf}
	\caption{
		\gammapy sub-package structure and data analysis workflow. The top row
        defines the different levels of data reduction, from lists of \gammaray-like
        events on the left (DL3), to high-level scientific products products
        (DL5) on the right. The direction of the data flow is illustrated with the
        gray arrows. The gray folder icons represent the different sub-packages
        in \gammapy and their names. Below each icon there is a list of the most
        important objects defined in the sub-package.
    }
	\label{fig:data_flow}
\end{figure*}
%
\label{sec:gammaray-data-analysis}
% This might not be the bast place compared to introduction
The data analysis process in \gammaray astronomy is usually split into two parts.
The first one deals with the data processing from camera measurement, calibration, event
reconstruction and selection to yield a list of reconstructed \gammaray event candidates.
This part of the data reduction sequence, sometimes referred to as low-level analysis,
is usually very specific to a given observation technique and even to a given instrument.

@@

To extract physically relevant information, such as the flux, spatial or spectral shape of one or more sources,
an analytical model is commonly adopted to describe the intensity of the radiation from gamma-ray sources
as a function of the energy, $E_{\rm true}$, and of the position in the FoV, $p_{\rm true}$:
\begin{equation}
    \Phi(p_{\rm true}, E_{\rm true}; \hat{\theta}), [\Phi] = \unit{TeV^{-1}.cm^{-2}.s^{-1}}
    \label{eq:general_source_model}
\end{equation}
where $\hat{\theta}$ is a set of model parameters that can be adjusted in a fit. To convert this analytical flux model
into a prediction on the number of gamma-ray events detected by an instrument, $N_{\rm pred}$,
with their estimated energy $E$ and position $p$, the model is convolved with the response function of the instrument.

@@

\subsection{gammapy.irf}
\label{ssec:gammapy-irf}
%
%
\begin{figure*}[t]
	\centering
	\includegraphics[width=1.\textwidth]{figures/irfs.pdf}
	\caption{
		Using \code{gammapy.irf} to read and plot instrument response functions.
		The left panel shows the effective area as a function of energy for
		the \cta, \hess, \magic, \hawc and \fermi instruments. The right panel shows
		the $68\%$ containment radius of the PSF as a function of energy for the \cta, \hess
		and \fermi instruments. The \cta IRFs are from the \textit{prod5} production. The \hess IRFs are from the DL3 DR1,
        using observation ID 033787. The \magic effective area is computed for a
        $20\,{\rm min}$ observation at the Crab Nebula coordinates. The
		\fermi IRFs use \textit{pass8} data and are also taken at the position of the Crab Nebula.
		The \hawc effective area is shown for the event classes $N_{Hit}=5 - 9$ as light gray
		lines along with the sum of all event classes as a black line. The \hawc IRFs are taken from
		the first public release of event data the \hawc collaboration. All IRFs do not correspond
		to the latest performance of the instruments, but still are representative of the 
		detector type and energy range. We exclusively relied on publicly available
		data provided by the collaborations. The data is also availabe in the
		\code{gammapy-data} repository.
    }
	\label{fig:irfs}
\end{figure*}
%

@@

\begin{table*}
    \begin{center}
        \begin{tabular}{lll}
         \hline
         Type & Description & Unit Equivalency \\
         \hline
         dnde & Differential flux at a given energy & \unit{TeV^{-1}.cm^{-2}.s^{-1}} \\
         e2dnde & Differential flux at a given energy  & \unit{TeV.cm^{-2}.s^{-1}} \\
         flux & Integrated flux in a given energy range & \unit{cm^{-2}.s^{-1}} \\
         eflux & Integrated energy flux in a given energy range & \unit{erg.cm^{-2}.s^{-1}}\\
         \hline
        \end{tabular}
    \end{center}
    \caption{Definition of the different SED types supported in \gammapy.}
    \label{tab:sed_types}
\end{table*}

@@

\subsection{Temporal Analysis}
\label{ssec:temporal-analysis}
A common use case in many astrophysical scenarios is to study the temporal
variability of a source. The most basic way to do this is to construct a
light curve, i.e., the flux of a source in each given time bin. In \gammapy, this
is done by using the \code{LightCurveEstimator} that fits the normalisation of a
source in each time (and optionally energy) band per observation, keeping constant
other parameters.
For custom time binning, an observation needs to be split into finer time bins using
the \code{Observation.select\_time} method. Figure~\ref{fig:hess_lightcurve_pks}
shows the light curve of the blazar PKS~2155-304 in different energy bands as
observed by the \hess telescope during an exceptional flare on the night of
July 29 - 30, 2006~\cite{2009A&A...502..749A}. The data are publicly available 
as a part of the HESS-DL3-DR1~\cite{HESS_DR1}. Each observation is first split into 10 min smaller
observations, and spectra extracted for each of these within a 0.11\textdegree~radius
around the source. A \code{PowerLawSpectralModel} is fit to all the datasets, leading
to a reconstructed index of $3.54 \pm 0.02$. With this adjusted spectral model
the \code{LightCurveEstimator} runs directly for two energy bands,
\qtyrange{0.5}{1.5}{TeV} and \qtyrange{1.5}{20}{TeV} respectively.
%
\begin{figure*}[t]
    \sidecaption
	\includegraphics[width=0.6666\textwidth]{figures/hess_lightcurve_pks.pdf}
	\caption{
        Binned light curves in two different energy bands for the source
        PKS~2155-304 in two energy bands (\qtyrange{0.5}{1.5}{TeV} and \qtyrange{1.5}{20}{TeV})
        as observed by the \hess telescopes in 2006. The coloured markers
        show the flux points in the different energy bands. The horizontal
        error illustrates the width of the time bin of 10~min. The vertical
        error bars show the associated asymmetrical flux errors. The marker
        is set to the center of the time bin.
    }
    \label{fig:hess_lightcurve_pks}
\end{figure*}
%
The obtained flux points can be analytically modelled using the available or
user-implemented temporal models. Alternatively, instead of  extracting a
light curve, it is also possible to directly fit temporal models to the reduced
datasets. By associating an appropriate \code{SkyModel}, consisting of both temporal
and spectral components, or using custom temporal models with spectroscopic
variability, to each dataset, a joint fit across the datasets will directly
return the best fit temporal and spectral parameters.

@@

The \fermi data is introduced at the data level DL4, and directly bundled in a
\code{MapDataset}. They have been prepared using the standard \textit{fermitools} \citep{Fermitools2019} and
selecting a region of $5^{\circ}\,\times\,4^{\circ}$ around the
position of the Crab Nebula, applying the same selection criteria of the 3FHL
catalog (7 years of data with energy from \qty{10}{GeV} to \qty{2}{TeV},
~\citealt{3FHL}).

The \magic data is included from the data level DL3. They consist of two
observations of $20\,{\rm min}$ each, chosen from the dataset used to estimate
the performance of the upgraded stereo system~\citep{magic_performance} and
already included in~\cite{joint_crab}. The observations were taken at small
zenith angles ($<30^{\circ}$) in wobble mode~\citep{fomin_1994}, with the
source sitting at an offset of $0.4^{\circ}$ from the FoV center. Their energy
range spans \qty{80}{GeV} to \qty{20}{TeV}. The data reduction for the 1D analysis
is applied, and the data are reduced to a \code{SpectrumDataset} before being fitted.

\hawc data are directly provided as flux points (DL5 data level) and are read
via Gammapy's \code{FluxPoints} class. They were estimated in ~\cite{hawc_crab_2019}
with $2.5\,{\rm years}$ of data and span an energy range \qty{300}{GeV} to \qty{300}{TeV}.

Combining the datasets in a \code{Datasets} list, \gammapy automatically generates
a likelihood including three different types of terms, two Poissonian likelihoods
for \fermi's \code{MapDataset} and MAGIC's \code{SpectrumDataset}, and a $\chi^2$
accounting for the \hawc flux points. For \fermi, a three-dimensional forward folding
of the sky model with the IRF is performed, in order to compute the predicted counts
in each sky-coordinate and energy bin. For \magic, a one-dimensional forward-folding
of the spectral model with the \irfs is performed to predict the counts in each estimated energy bin. A log
parabola is fitted to the almost five decades in energy \qty{10}{GeV} to \qty{300}{TeV}.
%% CN: is the LP formula specified elsewhere?
